---
title: "Uncertainty in cumulative proportions"
author: "Matthew Kay"
date: "2024-09-07"
output: github_document
---

## Setup

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
  fig.retina = 2,
  dev = "ragg_png"
)
```


```{r setup, message=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggdist)
theme_set(theme_ggdist())

set.seed(1234)
```

## Data

Create a data frame of:

- `.draw`: index of draw from the distribution in [1, 1000]
- `group`: a group in {a, b, c}
- `value`: value of the variable in {x, y, z}
- `proportion`: proportion in (0, 1) that sums to 1 within each `.draw` x `group`.

```{r}
df = brms::rdirichlet(3000, matrix(c(200, 100, 100, 100, 125, 150, 50, 200, 300), nrow = 3)) |>
  as.data.frame() |>
  setNames(c("x", "y", "z")) |>
  mutate(.draw = rep(1:1000, each = 3), group = rep(c("a", "b", "c"), 1000))  |>
  pivot_longer(c(x, y, z), names_to = "value", values_to = "proportion")

head(df, n = 12)
```

## Fuzzy bars

Fuzzy barplot (H/T Solomon Kurz for the idea; see [here](https://twitter.com/SolomonKurz/status/1372632774285348864?s=20) and [here](https://github.com/mjskay/ggdist/issues/109)):

```{r fuzzy_bars, fig.width = 3, fig.height = 4}
df |>
  ggplot(aes(y = proportion)) +
  geom_col(aes(x = .draw, fill = value), position = "stack", width = 1) +
  stat_pointinterval(
    aes(x = mean(.draw), group = value), 
    data = \(d) d |> 
      # need 1 - p because position_stack goes top to bottom
      # (or use position_stack(reverse = TRUE) above)
      mutate(proportion = 1 - cumsum(proportion), .by = c(.draw, group)) |> 
      # last proportion is pointless to plot because it is 0 with no uncertainty
      filter(value != "z")
  ) +
  scale_fill_brewer() +
  facet_grid(. ~ group, switch = "x") +
  coord_cartesian(expand = FALSE) +
  scale_x_continuous(breaks = NULL) +
  xlab("group")
```

## Gradient fuzzy bars

To recreate the above charts using a gradient instead of many smooshed-together bars,
we need to do some color mixing: we want to calculate how much of each color
should be at each cumulative proportion $p$ (each point along the y axis). For a given 
value (x, y, or z), this will be the probability that the cumulative proportion for
that value is greater than `p` times the probability that the cumulative proportion for
the preceding value is less than `p`:

```{r}
p = seq(0, 1 - .Machine$double.eps, length.out = 200)

df_curves = df |>
  mutate(cumulative_proportion = cumsum(proportion), .by = c(.draw, group)) |>
  reframe(
    `P(value)` = (1 - ecdf(cumulative_proportion)(p)) * ecdf(cumulative_proportion - proportion)(p),
    proportion = p,
    .by = c(group, value)
  )
```

We can check what these curves look like to make sure they're doing what we
want:

```{r color_curves}
df_curves |>
  ggplot(aes(x = proportion, y = `P(value)`, color = value)) +
  geom_line(linewidth = 1) +
  facet_grid(group ~ .) +
  scale_color_brewer()
```

Now we just have to mix the colors manually:

```{r gradient_bars, fig.width = 3, fig.height = 4}
library(farver)

# create and train the scale manually so we can use it to get the colors to mix
fill_scale = scale_fill_brewer(direction = -1)
fill_scale$train(c("x", "y", "z"))

df_curves |>
  # mix the colors manually in the OKLab color space (which is roughly perceptually 
  # uniform and which has better color mixing than Lab, especially for blues)
  summarise(
    color = encode_colour(`P(value)` %*% decode_colour(fill_scale$map(value), to = "oklab"), from = "oklab"),
    .by = c(group, proportion)
  ) |>
  ggplot() +
  # need this blank geom to create a fill legend, since we aren't actually using the fill
  # scale in the slab geometry (we already pre-computed the colors above)
  geom_blank(aes(fill = value), data = data.frame(value = c("x", "y", "z"))) +
  geom_slab(
    aes(y = proportion, x = group, fill = I(color), group = NA), 
    thickness = 1, fill_type = "auto", side = "both"
  ) +
  stat_pointinterval(
    aes(x = group, y = proportion, group = value), 
    data = df |> 
      # need 1 - p because position_stack goes top to bottom
      # (or use position_stack(reverse = TRUE) above)
      mutate(proportion = cumsum(proportion), .by = c(.draw, group)) |> 
      # last proportion is pointless to plot because it is 0 with no uncertainty
      filter(value != "z")
  ) +
  fill_scale +
  scale_y_continuous(limits = c(0, 1)) +
  coord_cartesian(expand = FALSE)
```





