---
title: "Diamond plot with data using geom_slab and geom_weave"
author: "Matthew Kay"
date: "2023-09-26"
output: github_document
---


```{r setup, warning = FALSE, message = FALSE}
library(palmerpenguins)
library(ggplot2)
library(ggdist)
library(dplyr)
library(tidyr)

theme_set(theme_ggdist())
knitr::opts_chunk$set(dev = "ragg_png", fig.retina = 2)
```

Let's fit a simple model for mean body mass of penguins based on this data:

```{r}
penguins_clean = filter(penguins, !is.na(sex))

penguins_clean |>
  ggplot(aes(x = body_mass_g, y = species, fill = sex)) +
  geom_weave(position = "dodge", linewidth = NA)
```

Here's a model for average body mass conditional on sex and species:

```{r}
m = lm(body_mass_g ~ species * sex, data = penguins_clean)
m
```

We can get conditional means and confidence intervals from the model:

```{r}
grid = penguins_clean |>
  modelr::data_grid(sex, species)

preds = cbind(grid, predict(m, newdata = grid, interval = "confidence"))
preds
```

To create the diamonds, we will reshape this data into a data frame with three
rows per `sex*species` combination, one for the point estimate, and one for each
end of the interval. We'll also add an `is_point` column that is `1` for the
point estimate and `0` for the interval endpoints.

```{r}
preds_long = preds |>
  pivot_longer(c(fit, lwr, upr)) |>
  mutate(is_point = as.integer(name == "fit"))

preds_long
```

We can use the `is_point` column to set the thickness of a slab geometry, so that
it has maximum thickness at the point estimate and zero thickness at the interval
endpoints. We also set `scale = 0.5` for `geom_slab()` so that the diamonds
are not too tall:

```{r}
penguins_clean |>
  ggplot(aes(x = body_mass_g, y = species, fill = sex)) +
  geom_weave(position = "dodge", linewidth = NA, alpha = 0.35) +
  geom_slab(
    aes(x = value, y = species, thickness = is_point), 
    position = "dodge", 
    data = preds_long, 
    side = "both", 
    scale = 0.35, 
    alpha = 0.65, 
    color = "gray25", 
    linewidth = 0.5
  ) +
  labs(
    title = "Diamond plot with data using geom_slab and geom_weave"
  )
```

A disadvantage of diamond plots like this is that more uncertain estimates become
visually larger. You could make the diamonds shrink their vertical height in proportion
to their uncertainty, to better mimic how densities (which have equal area) would look.
We can do this by calculating the width of the interval and making the thickness at the
point estimate inversely proportional to that width:

```{r}
preds_long = preds |>
  mutate(width = upr - lwr) |>
  pivot_longer(c(fit, lwr, upr)) |>
  mutate(is_point = as.integer(name == "fit"))

penguins_clean |>
  ggplot(aes(x = body_mass_g, y = species, fill = sex)) +
  geom_weave(position = "dodge", linewidth = NA, alpha = 0.35) +
  geom_slab(
    aes(x = value, y = species, thickness = is_point / width), 
    position = "dodge", 
    data = preds_long, 
    side = "both", 
    scale = 0.35, 
    alpha = 0.65, 
    color = "gray25", 
    linewidth = 0.5
  ) +
  labs(
    title = "Diamond plot with data using geom_slab and geom_weave"
  )
```
