---
title: "Hypothetical Outcome Plots (HOPs) for proportions"
output: github_document
bibliography: references.bib
link-citations: true
---

Here is a quick example of hypothetical outcome plots (HOPs) for proportions

## Setup

The following libraries are needed:

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(modelr)
library(rstanarm)
library(tidybayes)
library(gganimate)     # devtools::install_github("thomasp85/gganimate")
library(ggstance)
library(forcats)

theme_set(
  theme_light() +
  theme(panel.grid = element_blank())
)
```

## Data

We'll use some count data to illustrate:

```{r data}
df = data_frame(
  group = c("A","B","C"),
  count = c(100, 50, 30),
  proportion = count / sum(count)
)
df
```

## Model

One *possible* model for this data is as follows (N.B. this makes some strong assumptions about how the data were generated for the purposes of an example, your mileage may vary!):

```{r model, results = "hide"}
m = stan_glm(count ~ group, family = poisson, data = df)
```

## HOPs

Given that model, we could construct a HOPs [@Hullman2015; @Kale2018] bar chart illustrating the posterior distribution for the proportion in each group:

```{r hops, fig.width = 7, fig.height = 3, cache = TRUE}
n_hops = 100

p = df %>%
  data_grid(group) %>%
  add_fitted_draws(m, n = n_hops) %>%
  group_by(.draw) %>%
  mutate(proportion = .value / sum(.value)) %>%
  ggplot(aes(y = group, x = proportion)) +
  geom_colh(fill = "gray75") +
  geom_point(data = df, color = "red") +
  annotate("text", y = "C", x = .5, label = "Observed proportion", hjust = 0, color = "red") +
  annotate("segment", y = "C", yend = "C", x = .18, xend = .49, linetype = "dashed", color = "red") +
  annotate("text", y = 2.7, x = .5, label = "Posterior distribution for proportion", hjust = 0,
    color = "gray35") +
  annotate("segment", y = 2.7, yend = 2.7, x = .1, xend = .49, linetype = "dashed", color = "gray75") +
  xlim(0,1) +
  transition_states(.draw, transition_length = 1, state_length = 1)

animate(p, nframes = n_hops * 2, width = 600, height = 300)
```

## Quantile dotplots (static alternative)

If animation were not available (e.g. in a print medium), an alternative might be a quantile dotplot [@Kay2016; @Fernandes2018]:

```{r quantile-dotplots, fig.width = 7, fig.height = 4}
observed_label_data = data_frame(
  group = "C",
  label = "Observed proportion",
  x = .18, xend = .44, y = .55
)
dotplot_label_data = data_frame(
  group = "C",
  label = "100 approximately equally likely proportions",
  x = .18, xend = .44, y = .3
)


df %>%
  data_grid(group) %>%
  add_fitted_draws(m) %>%
  group_by(.draw) %>%
  mutate(proportion = .value / sum(.value)) %>%
  group_by(group) %>%
  do(data_frame(proportion = quantile(.$proportion, ppoints(100)))) %>%
  ggplot(aes(x = proportion)) +
  geom_dotplot(binwidth = .01, fill = "gray65", color = NA) +
  facet_grid(fct_rev(group) ~ .) +
  geom_text(aes(xend + .01, y, label = label), data = observed_label_data, hjust = 0, color = "red") +
  geom_segment(aes(x, y, xend = xend, yend = y), data = observed_label_data,
    linetype = "dashed", color = "red") +
  geom_vline(aes(xintercept = proportion), data = df, color = "red") +
  geom_text(aes(xend + .01, y, label = label), data = dotplot_label_data, hjust = 0, color = "gray35") +
  geom_segment(aes(x, y, xend = xend, yend = y), data = dotplot_label_data,
    linetype = "dashed", color = "gray65") +
  xlim(0,1) +
  ylab(NULL) +
  scale_y_continuous(breaks = NULL)
```


## References
