---
title: "Likert Sankey Dotplot"
author: "Matthew Kay"
date: "2024-10-14"
output: github_document
---

A quick riff on Matt Cooper and Zac Dempsey's [post on using Sankey diagrams to display Likert data](https://the-kids-biostats.github.io/posts/2024-10-14_likert_visualisation/likert_visualisations.html).

## Setup

```{r knitr_setup, include=FALSE}
knitr::opts_chunk$set(
  fig.retina = 2,
  dev = "ragg_png"
)
```


```{r setup, include=FALSE}
library(simstudy)
library(ggsankey)
library(ggalluvial)
library(likert)
library(patchwork)
library(gt)
library(gtsummary)
library(flextable)
library(dplyr)
library(tidyr)
library(ggdist)

theme_set(theme_ggdist())
```

## Data

Data from the original post:

```{r}
# dat_i is the intervention group
n <- 183 # Set the number of individuals
def <- defData(varname = "pre", formula = "1;5", dist = "uniformInt") # Pre values: uniformly distributed between 1 and 5
dat_i <- genData(n, def)
group_probs <- c(0.45, 0.45, 0.10)
set.seed(123) # For reproducibility
dat_i$grp <- sample(1:3, n, replace = TRUE, prob = group_probs)
dat_i$post <- dat_i$pre

dat_i$post[dat_i$grp == 2] <- pmin(dat_i$pre[dat_i$grp == 2] + (rbinom(sum(dat_i$grp == 2), 2, 0.2) + 1), 5) # Increase by 1, max 5
dat_i$post[dat_i$grp == 3] <- pmax(dat_i$pre[dat_i$grp == 3] - (rbinom(sum(dat_i$grp == 3), 2, 0.2) + 1), 1) # Decrease by 1, min 1
# dat_c is the control group
n <- 154 # Set the number of individuals
def <- defData(varname = "pre", formula = "1;5", dist = "uniformInt") # Pre values: uniformly distributed between 1 and 5
dat_c <- genData(n, def)
group_probs <- c(0.55, 0.25, 0.20)
set.seed(123) # For reproducibility
dat_c$grp <- sample(1:3, n, replace = TRUE, prob = group_probs)
dat_c$post <- dat_c$pre

dat_c$post[dat_c$grp == 2] <- pmin(dat_c$pre[dat_c$grp == 2] + (rbinom(sum(dat_c$grp == 2), 2, 0.2) + 1), 5) # Increase by 1, max 5
dat_c$post[dat_c$grp == 3] <- pmax(dat_c$pre[dat_c$grp == 3] - (rbinom(sum(dat_c$grp == 3), 2, 0.2) + 1), 1) # Decrease by 1, min 1

dat <- rbind(cbind(dat_i, group = "Intervention"), 
             cbind(dat_c, group = "Control")) %>% 
  mutate(post = as.integer(post)) %>% 
  select(-grp)

response_levels = c("Less than one cup/day", "About 1-2 cups/day", "About 3-4 cups/day", "About 5-6 cups/day","More than 6 cups/day")
dat <- dat %>% 
  mutate(
    pre_l = factor(pre, 1:5, labels = response_levels),
    post_l = factor(post, 1:5, labels = response_levels)
  )

head(dat, n = 8)
# head(dat, 5) %>%
  # thekids_table(colour = "Saffron", padding = 3)
```

Sankey plot from the original post:

```{r}
max_prop <- dat %>%
  select(id, group, pre, post) %>% 
  pivot_longer(cols = c(pre, post)) %>% 
  group_by(group, name) %>%
  count(value) %>%
  mutate(freq = n / sum(n)) %>% 
  .$freq %>%
  max

max_prop <- plyr::round_any(max_prop, 0.05, f = ceiling)

p_sankey <- dat %>% 
  filter(group == "Intervention") %>% 
  rename(Pre = pre,
         Post = post) %>% 
  make_long(Pre, Post) %>% 
  mutate(node = factor(node, levels = c(7,6,5,4,3,2,1)),
         next_node = factor(next_node, levels = c(7,6,5,4,3,2,1))) %>% 
  ggplot(aes(x = x, 
             next_x = next_x, 
             node = node, 
             next_node = next_node,
             fill = factor(node))) +
  geom_sankey(alpha = 0.7,
              node.color = 'black') +
  geom_sankey_label(aes(label = node), alpha = 0.75,
                    size = 3, color = "black", fill = "gray80") +
  scale_fill_viridis_d(option = "plasma", end = 0.85, guide = "none") +
  scale_x_discrete(expand = c(0.05,0.05)) +
  scale_y_continuous(breaks = NULL) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.title.y=element_blank(),
        axis.text.y=element_blank(),
        axis.ticks=element_blank(),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5)) +
  labs(title = "Pre-post",
       fill = "Pre",
       x = "")

p_sankey
```

## Dotplots + Sankey

Two thoughts occured to me when looking at the original post:

1. Because the histograms are displayed side-by-side, having the main axis be the x axis makes
it hard to compare the pre and post distributions. This could be fixed either by stacking the
histograms vertically or by rotating the histograms. If you do the latter, you can also mirror
one of them to do a more pyramid-style approach.

2. Dotplots / waffle plots would allow you to show individual datapoints, seeing proportions and
raw counts simultaneously, and also have a consistent coloring by the pre value.

Something like this:

```{r fig.width = 7, fig.height = 3.5}
p_pre = dat |>
  filter(group == "Intervention") |>
  ggplot() +
  # ordinarily I would not set binwidth manually here (and let geom_dots find the optimal binwidth),
  # but because we have to use geom_dots in two different charts we need to pick a shared binwidth
  # to keep things consistent. Can use geom_dots(verbose = TRUE) without a binwidth to find one.
  geom_dots(aes(group = NA, fill = pre_l, y = forcats::fct_rev(pre_l)), layout = "bar", side = "left", color = NA, binwidth = 0.17) +
  scale_fill_viridis_d(option = "plasma", end = 0.85, direction = -1, guide = "none") +
  scale_x_continuous(breaks = NULL) +
  labs(y = NULL, x = NULL)

p_post = dat |>
  filter(group == "Intervention") |>
  ggplot() +
  geom_dots(aes(group = NA, fill = pre_l, y = forcats::fct_rev(post_l), order = forcats::fct_rev(pre_l)), layout = "bar", side = "right", color = NA, binwidth = 0.17) +
  scale_fill_viridis_d(option = "plasma", end = 0.85, direction = -1, guide = "none") +
  scale_y_discrete(position = "right") +
  scale_x_continuous(breaks = NULL) +
  labs(y = NULL, x = NULL)

p_pre + p_sankey + p_post
```

