---
title: "Linear regression"
output:
  github_document:
    toc: true
---


## Setup

Libraries that might be of help:

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(magrittr)
library(ggplot2)
library(rstanarm)
library(modelr)
library(tidybayes)            # install from github
library(gganimate)            # install from github

theme_set(theme_light())
```



### Data

Let's generate some vaguely linear data:

```{r}
set.seed(123)
sample_size = 60
a = 1
b = 2
sigma = 2


df = data_frame(
  x = seq(1, 10, length.out = sample_size),
  y = rnorm(sample_size, a + b * x, sigma),
  group = rep(c("a", "b"), sample_size/2),
  x_centered = x - mean(x)
) 

df
```

```{r fig.width = 5, fig.height = 5}
df %>%
  ggplot(aes(x = x, y = y)) +
  geom_point() 
```

```{r}
m = stan_glm(y ~ x, data = df)
```

### Visualizations

#### "Half-eye" plot / interval + density

```{r fig.width = 5.5, fig.height = 5}
m %>%
  gather_draws(`(Intercept)`, x, sigma) %>%
  ggplot(aes(x = .value, y = .variable)) +
  geom_halfeyeh()
```

#### Quantile dotplots

```{r fig.width = 5, fig.height = 5}
m %>%
  gather_draws(`(Intercept)`, x, sigma) %>%
  do(data.frame(.value = quantile(.$.value, ppoints(100)))) %>%
  ggplot(aes(x = .value)) +
  geom_dotplot(binwidth = 0.04) +
  facet_grid(.variable ~ .)
```

#### HOPs of coefficients

```{r fig.width = 5, fig.height = 5}
m %>%
  gather_draws(`(Intercept)`, x, sigma) %>%
  filter(.draw %in% floor(seq_range(.draw, n = 100))) %>%
  ggplot(aes(x = .value, y = .variable)) +
  geom_point() +
  transition_manual(.draw)
```

Controlling framerate:

```{r fig.width = 5, fig.height = 5}
p = m %>%
  gather_draws(`(Intercept)`, x, sigma) %>%
  filter(.draw %in% floor(seq_range(.draw, n = 100))) %>%
  ggplot(aes(x = .value, y = .variable)) +
  geom_point() +
  transition_manual(.draw)

animate(p, fps = 5)
```

#### Fit lines with uncertainty bands

```{r fig.width = 5, fig.height = 5}
df %>%
  add_fitted_draws(m) %>%
  ggplot(aes(x = x, y = .value)) +
  stat_lineribbon(.width = c(.50, .80, .95), fill = "gray50", alpha = .25) +
  geom_point(aes(y = y), data = df)
```

#### Overplotted regression lines

```{r fig.width = 5, fig.height = 5}
df %>%
  add_fitted_draws(m, n = 100) %>%
  ggplot(aes(x = x, y = .value)) +
  geom_line(aes(group = .draw), alpha = 0.1, color = "red") +
  geom_point(aes(y = y), data = df)
```

#### Regression lines as animated HOPs

Basic example:

```{r fig.width = 5, fig.height = 5}
df %>%
  add_fitted_draws(m, n = 100) %>%
  ggplot(aes(x = x, y = .value)) +
  geom_line(color = "red") +
  geom_point(aes(y = y), data = df) +
  transition_manual(.draw)
```

Showing how to control speed:

```{r fig.width = 5, fig.height = 5}
p = df %>%
  add_fitted_draws(m, n = 100) %>%
  ggplot(aes(x = x, y = .value)) +
  geom_line(color = "red") +
  geom_point(aes(y = y), data = df) +
  transition_manual(.draw)

animate(p, fps = 3)
```



#### Posterior predictions with uncertainty bands

```{r fig.width = 5.5, fig.height = 5}
df %>%
  add_predicted_draws(m) %>%
  ggplot(aes(x = x, y = .prediction)) +
  stat_lineribbon() +
  scale_fill_brewer() +
  geom_point(aes(y = y), data = df)
```
