---
title: "Untitled"
author: "Matthew Kay"
date: "2/27/2022"
output: html_document
---

```{r setup}
library(ggplot2)
library(dplyr)
library(ggdist)
library(distributional)

theme_set(theme_ggdist())
```

This is a quick experiment / comparison of two different approaches to making gradients, particularly for `stat_lineribbon()`.

Our general goal is to use the density to make gradients:

```{r}
density_plot = tibble(
  x = seq(-4, 4, length.out = 201),
  d = dnorm(x)  
) %>%
  ggplot(aes(x, d)) +
  geom_line() +
  coord_cartesian(expand = FALSE, clip = "off")

density_plot
```

We want to find evenly-spaced values of densities at which to create intervals, so that
we can string a bunch of intervals together into a band.

We'll pick them to be in the interval `(0, max(dnorm(x)))`, where `max(dnorm(x))` = `dnorm(0)`:

```{r}
k = 20 # number of points
d = ppoints(k, a = 0) * dnorm(0)

density_plot +
  geom_hline(yintercept = d, alpha = 0.2)
```

To find the x positions corresponding to those density values, we
need the inverse of the Normal density function, which is (plus or minus):

```{r}
inv_dnorm = function(d) {
  sqrt(-2*log(d) - log(2) - log(pi))
}
```

Then we can find the corresponding x values:

```{r}
x = inv_dnorm(d)

density_plot +
  geom_hline(yintercept = d, alpha = 0.2) +
  geom_segment(aes(x = x, xend = x, y = 0, yend = d), data = data.frame(x, d), alpha = 0.5)
```

Now we can use these x values to determine the width of the corresponding quantile
intervals that would have those x values as their upper limits:

```{r}
w = 1 - (1 - pnorm(x))*2

density_plot +
  geom_hline(yintercept = d, alpha = 0.2) +
  stat_interval(aes(xdist = dist, y = 0, x = NULL), .width = w, data = data.frame(dist = dist_normal()), height = 0.05) +
  geom_segment(aes(x = x, xend = x, y = 0, yend = d), data = data.frame(x, d), alpha = 0.5) +
  scale_color_viridis_d(guide = "none")
```

Putting this together into a single function:

```{r}
inv_dnorm = function(d) {
  sqrt(-2*log(d) - log(2) - log(pi))
}

density_widths = function(n) {
  pp = ppoints(n, a = 0)
  d = pp * dnorm(0)
  x = inv_dnorm(d)
  2*pnorm(x) - 1
}
```

Or this alternative definition that simplifies the calculation a bit:

```{r}
density_widths = function(n) {
  pp = ppoints(n, a = 0)
  
  # this is equivalent to:
  # d = pp * dnorm(0)
  # x = inv_dnorm(d)
  x = sqrt(-2 * log(pp))
  
  2*pnorm(x) - 1
}

```

We want to color according the the density at the endpoints of the interval, hence the use of `mapply(density, xdist, xmin)` here:

```{r}
density_plot +
  geom_hline(yintercept = d, alpha = 0.2) +
  stat_interval(aes(xdist = dist, y = 0, x = NULL, color = stat(mapply(density, xdist, xmin))), .width = density_widths(100), data = data.frame(dist = dist_normal()), height = 0.05) +
  geom_segment(aes(x = x, xend = x, y = 0, yend = d), data = data.frame(x, d), alpha = 0.5) +
  scale_color_distiller()

```

Now we'll do the same with a lineribbon. Here's a distribution of lines:

```{r}
set.seed(1234)
n = 5000

df = tibble(
  .draw = 1:n,
  intercept = rnorm(n, 3, 1),
  slope = rnorm(n, 1, 0.25),
  x = list(-4:5),
  y = map2(intercept, slope, ~ .x + .y * -4:5)
) %>%
  unnest(c(x, y))
```

We can do the classic lineribbon gradient by giving a large number of values of `ppoints()`, but the
colors in this case will map onto quantile interval probabilities (essentially, a transformation of the CDF)
whereas we want the color to map onto densities:

```{r}
df %>%
  ggplot(aes(x, y)) +
  stat_ribbon(aes(fill_ramp = stat(.width)), .width = ppoints(100), fill = "#2171b5") +
  scale_fill_ramp_continuous(range = c(1, 0), guide = "none") +
  coord_cartesian(ylim = c(-5, 15))
```

Notice how the gradient looks sharp in the middle even without the median line drawn.
This is because the CDF will also have a sharp peak at the median.

We can instead use the Normal approximation to the density at a given width to encode
the approximate density in color:

```{r}
# parameterize by `dist` in case we wanted to use a different distribution for the approximation
density_for_width = function(w, dist = "norm", ...) {
  quantile_fun = match.fun(paste0("q", dist))
  density_fun = match.fun(paste0("d", dist))
  (density_fun(quantile_fun(0.5 - w/2, ...), ...) + density_fun(quantile_fun(0.5 + w/2, ...), ...))/2
}


df %>%
  ggplot(aes(x, y)) +
  stat_ribbon(aes(fill_ramp = stat(density_for_width(.width))), .width = density_widths(100), fill = "#2171b5") +
  scale_fill_ramp_continuous(guide = "none") +
  coord_cartesian(ylim = c(-5, 15))
```

This should better match what we would get if we used `stat_gradientinterval()`, which
directly encodes the density on color (except for the difference that gradientinterval
does not map the max density conditional on x onto the same color, and of course the
discontinuities between x values):

```{r}
df %>%
  ggplot(aes(x, y)) +
  stat_gradientinterval(fill = "#2171b5", color = NA, fill_type = "gradient", scale = 1) +
  coord_cartesian(ylim = c(-5, 15))
```
