---
title: "Multivariate regression"
output:
  github_document:
    toc: true
---


## Setup

Libraries that might be of help:

```{r setup, message = FALSE, warning = FALSE}
library(tidyverse)
library(magrittr)
library(ggplot2)
library(rstan)
library(brms)
library(modelr)
library(tidybayes)
library(ggridges)
library(patchwork)  # devtools::install_github("thomasp85/patchwork")

theme_set(theme_light())
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```



### Data

```{r}
set.seed(1234)

df =  data_frame(
  y1 = rnorm(20),
  y2 = rnorm(20, y1),
  y3 = rnorm(20, -y1)
)
```

### Data plot

```{r}
df %>%
  gather(.variable, .value) %>%
  gather_pairs(.variable, .value) %>%
  ggplot(aes(.x, .y)) +
  geom_point() +
  facet_grid(.row ~ .col)
```

### Model

```{r}
m = brm(cbind(y1, y2, y3) ~ 1, data = df)
```

### Correlations from the model

A plot of the `rescor` coefficients from the model:

```{r}
m %>%
  gather_draws(`rescor.*`, regex = TRUE) %>%
  separate(.variable, c(".rescor", ".row", ".col"), sep = "__") %>%
  ggplot(aes(x = .value, y = 0)) +
  geom_halfeyeh() +
  xlim(c(-1, 1)) +
  xlab("rescor") +
  ylab(NULL) +
  facet_grid(.row ~ .col)
```

### Altogether

I'm not sure I like this (we're kind of streching the limits of `facet_grid` here...) but if you absolutely must have a combined plot, this sort of thing could work...

```{r}
correlations = m %>%
  gather_draws(`rescor.*`, regex = TRUE) %>%
  separate(.variable, c(".rescor", ".row", ".col"), sep = "__")


df %>%
  gather(.variable, .value) %>%
  gather_pairs(.variable, .value) %>%
  ggplot(aes(.x, .y)) +
  
  # scatterplots
  geom_point() +

  # correlations
  geom_halfeyeh(aes(x = .value, y = 0), data = correlations) +
  geom_vline(aes(xintercept = x), data = correlations %>% data_grid(nesting(.row, .col), x = c(-1, 0, 1))) +

  facet_grid(.row ~ .col)
```

### Or side-by-side

Actually, it occurs to me that the traditional "flipped on the axis" double-scatterplot-matrix can be hard to read, because it is hard to mentally do the diagonal-mirroring operation to figure out which cell on one side goes with the other. I find it easier to just map from the same cell in one matrix onto another, which suggests something like this might be better:

```{r, fig.width = 8, fig.height = 4}
data_plot = df %>%
  gather(.variable, .value) %>%
  gather_pairs(.variable, .value) %>%
  ggplot(aes(.x, .y)) +
  geom_point() +
  facet_grid(.row ~ .col)

rescor_plot = m %>%
  gather_draws(`rescor.*`, regex = TRUE) %>%
  separate(.variable, c(".rescor", ".col", ".row"), sep = "__") %>%
  ggplot(aes(x = .value, y = 0)) +
  geom_halfeyeh() +
  xlim(c(-1, 1)) +
  xlab("rescor") +
  ylab(NULL) +
  facet_grid(.row ~ .col)

data_plot + rescor_plot
```

### More heatmap-y

Some other things possibly worth improving:

- adding a color encoding back in for that high-level gist
- making "up" be positive correlation and "down" be negative
- 0 line

```{r, fig.width = 8, fig.height = 4, warning = FALSE, message = FALSE}
rescor_plot_heat = m %>%
  gather_draws(`rescor.*`, regex = TRUE) %>%
  separate(.variable, c(".rescor", ".col", ".row"), sep = "__") %>%
  ggplot(aes(x = .value, y = 0)) +
  geom_density_ridges_gradient(aes(fill = stat(x)), color = NA) +
  stat_pointintervalh() +
  xlim(c(-1, 1)) +
  xlab("rescor") +
  ylab(NULL) +
  scale_y_continuous(breaks = NULL) +
  scale_fill_distiller(type = "div", palette = "RdBu", direction = 1, limits = c(-1, 1), guide = FALSE) +
  geom_vline(xintercept = 0, color = "gray65", linetype = "dashed") +
  coord_flip() +
  facet_grid(.row ~ .col)

data_plot + rescor_plot_heat
```
