---
title: "Multivariate measurement error model in brms"
author: "Matthew Kay"
date: "2023-02-20"
output: html_document
---

```{r}
library(tidyverse)
library(brms)
library(posterior)
library(ggdist)

theme_set(theme_ggdist())
options(mc.cores = parallel::detectCores())
```

# Generate data

We'll generate some bivariate Normal data for two variables, `x1` and `x2`, 
with a known correlation (`true_cor`) and a known bias (`true_x2_bias`).

We'll also generate "observations" of these variables, `x1_obs` and `x2_obs`,
with measurement error `x1_sd` and `x2_sd`. i.e. we're assuming we can't observe
`x1` or `x2` directly (so they won't be in our model).

```{r}
set.seed(1234)

# reparameterize gamma dist in terms of mean and precision
# this will allow us to generate distributions of measurement error SDs
rgamma_mean_prec = function(n, mean, prec) rgamma(n, prec * mean / (mean + 1), prec / (mean + 1))

true_x2_bias = 0.1
true_cor = 0.6

df = MASS::mvrnorm(
  100, 
  mu = c(x1 = 0, x2 = true_x2_bias), 
  Sigma = matrix(c(1,true_cor,true_cor,1), nrow = 2)
) |>
  as_tibble() |>
  mutate(
    x1_sd = rgamma_mean_prec(n(), 0.5, 500),
    x1_obs = rnorm(n(), x1, x1_sd),
    # give x2 higher measurement error on average compared to x1, in analogy
    # to how adaptive will have higher measurement error
    x2_sd = rgamma_mean_prec(n(), 0.8, 500),
    x2_obs = rnorm(n(), x2, x2_sd)
  )
```

The first couple of rows of the data:

```{r}
head(df)
```

Plotted, with (unobservable) `x1` and `x2` as black squares and observed `x1_obs` and
`x2_obs` as blue dots +/- 2*SD:

```{r}
df |>
  ggplot(aes(x1, x2)) +
  geom_point(aes(x1_obs, x2_obs), color = "blue", alpha = 0.35) +
  geom_linerange(aes(x = x1_obs, y = x2_obs, ymin = x2_obs - 2*x2_sd, ymax = x2_obs + 2*x2_sd), color = "blue", alpha = 0.35) +
  geom_linerange(aes(x = x1_obs, y = x2_obs, xmin = x1_obs - 2*x1_sd, xmax = x1_obs + 2*x1_sd), color = "blue", alpha = 0.35) +
  geom_point(alpha = 0.75, shape = "square") +
  coord_fixed()
```

A model of this data generating process is something like this:

$$
\begin{align}
x_1^\textrm{obs} &\sim \operatorname{Normal}\left(x_1, x_1^\textrm{sd}\right)\\
x_2^\textrm{obs} &\sim \operatorname{Normal}\left(x_2, x_2^\textrm{sd}\right)\\
\begin{bmatrix}x_1 \\ x_2 \end{bmatrix} &\sim \operatorname{MVN}\left(
  \begin{bmatrix} \mu_1 \\ \mu_2 \end{bmatrix},
  \begin{bmatrix} \sigma_1^2 & \sigma_1\rho\sigma_2 \\ \sigma_1\rho\sigma_2 & \sigma_2^2  \end{bmatrix}
\right)
\end{align}
$$

Where $x_1$ and $x_2$ are unobserved, and $\rho$ (`true_cor`) is the correlation between $x_1$ and $x_2$.
This looks a bit more complicated than our data generating process above, but that's because
we fixed $\mu_1 = 0$ so that $\mu_2$ could be the bias of $x_2$ relative to $x_1$ (`true_x2_bias`). 
We also fixed both $\sigma$s to be 1, which simplified the data generating process to this:

$$
\begin{align}
\begin{bmatrix}x_1 \\ x_2 \end{bmatrix} &\sim \operatorname{MVN}\left(
  \begin{bmatrix} 0 \\ \mu_2 \end{bmatrix},
  \begin{bmatrix} 1 & \rho \\ \rho & 1  \end{bmatrix}
\right)
\end{align}
$$

However, we cannot assume those simplifications hold when fitting the model, so 
we estimate all those parameters from the data. The corresponding model written
in brms syntax is:

```{r}
m_formula = 
  bf(x1_obs | mi(x1_sd) ~ 1) +
  bf(x2_obs | mi(x2_sd) ~ 1) +
  set_rescor(TRUE)
```

For example, `x1_obs | mi(x1_sd)` defines a measurement error portion of the model, like 
$x_1^\textrm{obs} \sim \operatorname{Normal}\left(x_1, x_1^\textrm{sd}\right)$, and the
combination of the two formulas with `+ set_rescor(TRUE)` tells brms to fit a multivariate
model and estimate the residual correlation, defining the multivariate normal ($\operatorname{MVN}$)
portion of the model.

Let's check on priors:

```{r}
get_prior(m_formula, data = df)
```

I'll set some simple priors on this for the purposes of the example (should not
just use these directly on the IRT stuff...)

```{r}
m_prior = c(
  prior(normal(0, 1), class = Intercept, resp = x1obs),
  prior(normal(0, 1), class = Intercept, resp = x2obs),
  prior(lkj(2), class = rescor),
  prior(normal(1, 1), class = sigma, resp = x1obs),
  prior(normal(1, 1), class = sigma, resp = x2obs)
)
```

And fit the model:

```{r}
m = brm(
  m_formula,
  prior = m_prior,
  backend = "cmdstanr",
  data = df
)
```

Results look like this:

```{r}
m
```

This gives us an estimate of the bias:

```{r}
m |>
  as_draws_df() |>
  mutate(bias = b_x2obs_Intercept - b_x1obs_Intercept) |>
  ggplot(aes(x = bias)) +
  stat_halfeye() +
  geom_vline(xintercept = true_x2_bias)
```

And an estimate of the correlation:

```{r}
m |>
  as_draws_df() |>
  ggplot(aes(x = rescor__x1obs__x2obs)) +
  stat_halfeye() +
  geom_vline(xintercept = true_cor)
```
