---
title: "Uncertainty with Palmer Penguins"
format: html
editor: visual
---

## Setup

The following libraries are needed:

```{r}
library(palmerpenguins)  # the data
library(ggplot2)         # for plotting
library(dplyr)           # for data cleaning
library(ggdist)          # for visualizing distributions and uncertainty
library(distributional)  # for creating distributions, e.g. dist_normal()
library(broom)           # for tidying model output: augment()

# a nicer ggplot theme...
theme_set(theme_ggdist())
```


## The data --- Penguins!

We'll be using the Palmer Penguins dataset, which you can learn more about [here](https://allisonhorst.github.io/palmerpenguins/).

It contains data on several species of penguins who live in Antarctica, and 
includes data on their body mass, sex, and dimensions of their bills and
flippers:

```{r}
glimpse(penguins)
```

For example, here is a simple scatter plot of body mass by sex and species:

```{r}
penguins |>
  ggplot(aes(x = body_mass_g, y = species, color = sex)) +
  geom_point()
```

As you can see, it's a bit hard to see what's going on just using a scatterplot, because
of the overlap between points.

Let's use `ggdist::geom_dots()` with `position_dodge()`
to make it easier to see...

```{r}
penguins |>
  ggplot(aes(x = body_mass_g, y = species, fill = sex)) +
  geom_dots(position = "dodge", linewidth = 0)
```

We can see a few missing data values in `sex` that we will remove for now:

```{r}
penguins_clean = penguins |>
  filter(!is.na(sex))

penguins_clean |>
  ggplot(aes(x = body_mass_g, y = species, fill = sex)) +
  geom_dots(position = "dodge", linewidth = 0)
```


## Adding a model and showing uncertainty

Say we want an estimate of the average mass of a penguin, given its species 
and sex. We could fit a linear model to estimate this:

```{r}
m = lm(body_mass_g ~ species * sex, data = penguins_clean)
m
```

Then, we can use the `broom` package to get estimates of the body mass for each
combination of species and sex in the data. To do that, we'll first create a 
prediction grid containing all combinations of the predictors we want to
estimate mean body mass for:

```{r}
grid = penguins_clean |>
  select(species, sex) |>
  unique()

grid
```

Then, we'll use `broom::augment()` to get means (`.fitted`) and standard errors (`.se.fit`) for each
combination of species and sex:

```{r}
predictions = augment(m, newdata = grid, se_fit = TRUE)
predictions
```

Finally, since we'll be using a Student t distribution to calculate intervals,
we need the degrees of freedom of the fit:

```{r}
dof = df.residual(m)
dof
```

Now, we can generate a visulization of uncertainty in the mean body mass conditional
on sex and speciies:

```{r}
predictions |>
  ggplot(aes(y = species, fill = sex)) +
  stat_halfeye(aes(xdist = dist_student_t(dof, .fitted, .se.fit)), scale = 0.5) 
```

We can combine the estimate of the mean with the raw data to see how they line up:

```{r}
predictions |>
  ggplot(aes(y = species, fill = sex)) +
  geom_swarm(aes(x = body_mass_g), data = penguins_clean, position = "dodge", linewidth = 0, height = 0.7, alpha = 0.5) +
  stat_pointinterval(aes(xdist = dist_student_t(dof, .fitted, .se.fit)), position = "dodge", scale = 0.5, height = 0.7)
```
