---
title: "Dotplots of Wee People"
author: "Matthew Kay"
date: '2022-04-27'
output: github_document
---

```{r include=FALSE}
knitr::opts_chunk$set(
  dev = "png",
  dev.args = list(png = list(type = "cairo")),
  fig.retina = 2
)
```


## Introduction

This demo shows the use of the [Wee People font](https://github.com/propublica/weepeople)
from ProPublica with `geom_dots()` from [ggdist](https:://mjskay.github.io/ggdist). The idea
of using Wee People with dotplots comes from @gdbassett, and the code 
is based on [@hrbrmstr's example code](https://twitter.com/hrbrmstr/status/1363241620527808514?s=20&t=Dz0e9sEqPYg8d5J5_37fDQ)
for using custom fonts in ggplot.

## Setup

The dev version of `ggdist` is needed to run this example. Install it with:

```r
remotes::install_github("mjskay/ggdist")
```

These libraries are needed:

```{r setup}
library(systemfonts)
library(ggplot2)
library(ggdist)

theme_set(theme_ggdist())
```

```{r}
knitr::opts_chunk$set(dev = "png", dev.args = list(png = list(type = "cairo")))
```

## Font setup

Download the Wee People font:

```{r}
download.file(
  url = "https://github.com/propublica/weepeople/raw/master/weepeople.ttf",
  destfile = "weepeople.ttf"
)
```

And register it so we can use it with `family = "weepeople"`:

```{r}
register_font(
  name = "weepeople",
  plain = "weepeople.ttf"
)
```


## Wee People Dotplots

An example with raw data using `geom_dots()`:

```{r}
set.seed(1234)

df = data.frame(
  x = qnorm(ppoints(100), 1:2),
  set = c("a", "b"),
  icon = factor(sample(52, 100, replace = TRUE))
) 

# the lower and upper case letters in the Wee People font are people:
people = c(letters, toupper(letters))

df |>
  ggplot(aes(x = x, y = set, group = set, shape = icon)) + 
  geom_dots(family = "weepeople") + 
  scale_shape_manual(values = people, guide = "none")
```

Notice how the people are spaced out horizontally. This is because compared to 
normal dots, which have a square aspect ratio, they are quite tall. With `geom_dots()`,
`stackratio` determines vertical stacking and `dotsize` horizontal spacing, so if we
increase `dotsize` we can make them be spaced closer together (fortunately the
automatic bin width selection algorithm will account for this and adjust binwidth
accordingly):

```{r}
df |>
  ggplot(aes(x = x, y = set, group = set, shape = icon)) + 
  geom_dots(family = "weepeople", dotsize = 2.4) + 
  scale_shape_manual(values = people, guide = "none")
```

We can also make it feel a bit more organic using the `"swarm"` layout:

```{r}
df |>
  ggplot(aes(x = x, y = set, group = set, shape = icon, color = x > 0)) + 
  geom_dots(family = "weepeople", dotsize = 2.4, layout = "swarm")+#, binwidth = 0.145) + 
  scale_shape_manual(values = people, guide = "none") +
  scale_color_brewer(palette = "Set1", guide = "none")
```

For completeness, an example with an analytical distribution using `stat_dots()`:

```{r}
data.frame(x = distributional::dist_normal(0,1)) |>
  ggplot(aes(xdist = x, shape = stat(factor(1:50)))) + 
  stat_dots(quantiles = 50, family = "weepeople", dotsize = 2.4) + 
  scale_shape_manual(values = people, guide = "none")
```

## Bonus

Bonus: emoji dotplots using {ragg}

```{r}
ragg::agg_png("test.png", width = 3000, height = 2000, res = 300)
print(df |>
  ggplot(aes(x = x, y = set, group = set, color = x > 0)) + 
  geom_dots(aes(shape = x > 0), layout = "weave", dotsize = 0.9, stackratio = 1.1, fill = NA, position = position_nudge(x = 0.07)) + 
  # geom_dots(layout = "swarm", dotsize = 0.9, stackratio = 1.1, fill = NA) +
  scale_shape_manual(values = c("💩", "😺"), guide = "none") +
  scale_color_brewer(palette = "Set1", guide = "none")) +
  geom_vline(xintercept = 0)
dev.off()
```

